---
title: "Biomarker prioritization and power analysis"
author: "Ivan Krukov"
date: "December 15, 2020"
output:
  beamer_presentation: 
      theme: "Pittsburgh"
      fonttheme: "structurebold"
  ioslides_presentation: default
---

```{r setup, include=FALSE}
library(lme4)
library(lmerTest)
library(tidyverse)
source("simulations.R")
ggplot2::theme_set(theme_minimal(base_size = 18))
```

## Problem statement

- Treatment and placebo groups
- 10 patients each
- Baseline, then 5 days of follow-up
- 4000 biomarkers measured each day

## Potential approaches

- Repeated measures ANOVA
  - Appropriate, but no effect indication
- Linear mixed model
  - Effect size inference, significance testing
- Bayesian linear model
  - Appropriate with limited data, but priors are hard to choose
- Machine learning
  - Performant, but hard to interpret
  
# Analysis pipeline
  
## Data pre-processing

- Log-transformation
  - Necessary if biomarker varies over degrees of magnitude
- Z-score standardization
  - 0-centered 
  - Units of standard deviations

## Simulated dataset

```{r sim_data, echo = F}
set.seed(100)
param <- list(eps_sd = .5, base = 0,
                trt = list(eff = .1, base_sd = 1, eff_sd = .025),
                plc = list(eff = 0,  base_sd = 1, eff_sd = 0))

df <- sim_experiment(10, 10, 5, param)
df_plot <- ggplot(df, aes(t, y)) + 
  geom_line(aes(group = id)) + 
  facet_wrap(~ group) +
  labs(x = "Time, days", y = "Effect, SD")
df_plot
```

## Mixed-effects model

- Fixed effects
  - On average, baseline is zero (standardized)
  - Biomarker activity increases over time (only with treatment)

- Random effects
  - Baseline is different for each patient
  - Individual response to drug is different
  
## Inference

```{r, message=F, echo=F}
model <- lmer(y ~ t:group + (t | id), data = df)

coeff_table <- summary(model)$coefficients %>% 
  as.data.frame() %>% 
  round(3) %>% 
  rownames_to_column("Fixed effect") %>%
  rename(`p value` = `Pr(>|t|)`)

coeff_table %>% knitr::kable(caption = "Inferred effects")
```

## Inference results

```{r, echo=F}
fe <- fixef(model)
estimates <- tibble(
  group = c('Placebo', 'Treatment'), 
  base = c(fe[1], fe[1]),
  eff = c(fe[2], fe[3]))
df$pred <- predict(model)
df_plot
```

## Inference results

```{r, echo=F}
ggplot(df, aes(t, y)) + 
  geom_line(aes(group = id)) + 
  geom_line(aes(y = pred, group = id), linetype = 'dashed') +
  geom_abline(
    aes(slope = eff, intercept = base, group = group), 
  data = estimates, color = 'blue', size = 1.5) +
  facet_wrap(~ group) +
  labs(x = "Time, days", y = "Effect, SD")
```

## Multiple testing correction

- Benjamini-Hotchberg False discovery rate
  - Controls for the false positives in the entire experiment
  - Less stringent than Bonferonni correction

# Improving the design

## Power of the test

- More data!
- How many patients do we need per group?
- How many days of follow-up do we need?

## Participant number

```{r, echo=F, message=F}
participant_power <- read_csv('cached/participant_power.csv')
critical_point <- participant_power %>% 
  filter(mean > 0.9) %>% 
  pull(participants) %>% .[1]

participant_power %>%
  ggplot(aes(x = participants)) + 
  geom_ribbon(aes(ymin = pmax(0, mean - sd), ymax = pmin(1, mean + sd)), fill = 'gray90') +
  geom_line(aes(y = mean)) +
  geom_hline(yintercept = 0.9, color = 'blue', linetype = 'dashed') +
  geom_text(y = 0.85, x = critical_point, label = paste0(">", critical_point), color = 'black', size = 5) +
  labs(x = "Number of participants", y = "Power")
```

## Days of follow-up

```{r, echo=F, message=F}
day_power <- read_csv('cached/day_power.csv')
critical_point <- day_power %>% 
  filter(mean > 0.9) %>% 
  pull(days) %>% .[1]

day_power %>%
  ggplot(aes(x = days)) + 
  geom_ribbon(aes(ymin = pmax(0, mean - sd), ymax = pmin(1, mean + sd)), fill = 'gray90') +
  geom_line(aes(y = mean)) +
  geom_hline(yintercept = 0.9, color = 'blue', linetype = 'dashed') +
  geom_text(y = 0.85, x = critical_point, label = paste0(">", critical_point), color = 'black', size = 5) +
  labs(x = "Days of follow-up", y = "Power")
```

## Further considerations

- Is the sample truly representative?
- Does the effect increase indefinitively with time?

- Smaller placebo group - 20 participant may be sufficient
- Effect of the drug concentration

Code available at:

[github.com/ivan-krukov/biomarker-power](https://github.com/ivan-krukov/biomarker-power)

# Questions?

## Assumptions about effect

- With treatment, biomarker level changes 0.05 SD per day
 - 2% improvement per day
 - 93rd percentile after 30 days of treatment

```{r}
param <- list(
  eps_sd = .5, base = 0,
  trt = list(eff = .05, base_sd = 1, eff_sd = .025),
  plc = list(eff = 0,  base_sd = 1, eff_sd = 0))
```

